\documentclass{article}
\usepackage{graphicx}
\usepackage[dot, autosize, outputdir="graphfiles/"]{dot2texi}
\usepackage{tikz}
\usepackage{amsmath, mathtools}
\usetikzlibrary{shapes}
\title{W3}
\author{Zeerak Waseem - csp265}
\date{8/12/2013}
\begin{document}
\maketitle
\newpage

\section{Exercise 1}

\subsection{a}
When f(a, b, c) returns, it returns \(7 + 3 + 12 = 22\). Thus when call-by-value is used, \(r = 22\) \(x\) and \(y\) remain unchanged in the calling function. The reason is that call-by-value creates a copy of the variables to be used in the called function, thus leaving the variables unchanged in the calling function.
\subsection{b}
When call-by-reference is used, the variables in the calling function are modified, thus \(x\) to be first \(7\) then subsequently changing \(x\) again to be \(12\). \(y\) is changed to \(3\).
\subsection{c}
When call-by-value-result is used, the variables are only changed when the called function returns, leaving the values in the calling function unchanged (as such, \(x\) would only change once rather than twice, as with call-by-reference. Thus \(x\) and \(y\) would be modified to \(12\) and \(3\) respectively.

\section{Exercise 2}
\subsection{a}
It would print 5 twice as g() has no access to the local \(x's\), but only to the global.
\subsection{b}
It would print 4 and 9, as g() does have access to the local variables.
\subsection{c}
By editing in the \texttt{TpInterpret.sml} somewhere between line 226 and 285. I'm unsure where.

\section{Exercise 3}
\subsection{a}
\subsection{b}
\subsection{c}
\end{document}
